---
title: "Getting started with Data"
author: "R Carpentry"
date: '2022-08-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lesson plan:   
 - Load external data from a .csv file into R  
 - See how R interprets these data in the R environment as a dataframe  
 - Examine the data in the form of a dataframe in R  
 - Create and change the types of data in a dataframe, including factors and strings  
 - Reorder and rename factors  
 - try plotting some data!  
 - knit the file!  
  
  
  
Downloading data and looking at dataframes.
```{r download data}
# If you didn't get the package installed yet, you can type 
# install.packages("tidyverse")

# load the library tidyverse
library(tidyverse)

# set your working directory (i.e. what folder you're working in)

# create an object that will be our dataset

# look at the first few lines of the dataset

# print the dataset

# look at the last few lines of the dataset

# pull the data up in a new window

# usually good to comment this command out



# What does a horizontal row in this dataframe represent?

# What does a vertical column in this dataframe represent?






# what is the structure of our data? -----


# How many observations do we have?


# How many variables do we have collected from each observation?





```

What is factored data?

```{r factored data}

# What does it mean for a variable to be a factor?


# Let's change the assignment of this column from a character to a factor

# We can see what this function did by using the summary() function.





# Let's create a factored vector to explore more about vectors ----
# our vector, series will include factors: one, two, three, two


# what's the order of these factors?


# how many levels are there? use the nlevels() function


# we want to order these levels
# using the levels argument


# call the levels of series now

```

Let's try converting our variables.

```{r converting factors}

# Converting factors -----

# You can also convert factors back to characters or numeric values 





# Converting a factor to a numeric vector is a little more tricky
# first we'll try with the factor() function

# If we use the as.numeric() function on its own:


# To fix:
# we need to first make the numbers into characters and THEN convert to numeric


# Or, we can use the levels() function, where we make the levels of year_fct
# numeric, and then use that to apply to the year_fct vector. 

```

Let's plot some data!

```{r plotting based on factors}

# Plotting based on factors --------------------------------------------------------

# When you have factored data, you can plot your data based on those factors to see,
# for example, a count of the number of observations in each plot type



# Maybe we might want to know if animals are bigger in one of the treatment groups


```