---
title: "Getting started with Data"
author: "R Carpentry"
date: '2022-08-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lesson plan:   
 - Load external data from a .csv file into R  
 - See how R interprets these data in the R environment as a dataframe  
 - Examine the data in the form of a dataframe in R  
 - Create and change the types of data in a dataframe, including factors and strings  
 - Reorder and rename factors  
 - try plotting some data!  
 - knit the file!  
  
  
  
Downloading data and looking at dataframes.
```{r download data}
# If you didn't get the package installed yet, you can type 
# install.packages("tidyverse")

# load the library tidyverse
library(tidyverse)

# set your working directory (i.e. what folder you're working in)
#Error message: cannot change working directory

# create an object that will be our dataset
surveys <- read_csv("surveys.csv")

# look at the first few lines of the dataset
head(surveys)

# print the dataset
print(surveys)

# look at the last few lines of the dataset
tail(surveys)

# pull the data up in a new window
#view(surveys)

# usually good to comment this command out



# What does a horizontal row in this dataframe represent?
  # An observation of a specific animal

# What does a vertical column in this dataframe represent?
  # Some attribute of those data, including the data of observation, the species and size of the animal, and the type of plot they were observed in in our experiment





# what is the structure of our data? -----
str(surveys)

# How many observations do we have?
  # 34,786 observations

# How many variables do we have collected from each observation?
  # 13 variables
# what the first few values for each of these variables are and also what kind of variable they are: eg numeric vs. character


```

What is factored data?

```{r factored data}

# What does it mean for a variable to be a factor?
# Factored data means that a column in your dataframe that might have numeric or character values may actually have a "hidden meaning" in which those numbers or characters actually represent groups or categories of data

# Let's change the assignment of this column from a character to a factor
surveys$plot_id <- factor(surveys$plot_id)

# We can see what this function did by using the summary() function.
summary(surveys$plot_id)




# Let's create a factored vector to explore more about vectors ----
# our vector, series will include factors: one, two, three, two
series <- factor(c("one", "two", "three", "two"))

# what's the order of these factors?
levels(series) # alphabetical

# how many levels are there? use the nlevels() function
nlevels(series)

# we want to order these levels
# using the levels argument
series <- factor(series, levels = c("one", "two", "three"))

# call the levels of series now
levels(series)

```

Let's try converting our variables.

```{r converting factors}

# Converting factors -----

# You can also convert factors back to characters or numeric values 

as.character(series) # this can be helpful when we plot

class(series)

series <- as.character(series)

class(series)

# Converting a factor to a numeric vector is a little more tricky
# first we'll try with the factor() function
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

# If we use the as.numeric() function on its own:
as.numeric(year_fct) # returns just the level of the factor each number corresponds to

# To fix:
# we need to first make the numbers into characters and THEN convert to numeric
as.numeric(as.character(year_fct))

# Or, we can use the levels() function, where we make the levels of year_fct
# numeric, and then use that to apply to the year_fct vector. 
as.numeric(levels(year_fct))[year_fct]

```

Let's plot some data!

```{r plotting based on factors}

# Plotting based on factors --------------------------------------------------------

# When you have factored data, you can plot your data based on those factors to see,
# for example, a count of the number of observations in each plot type
plot(surveys$plot_id)


# Maybe we might want to know if animals are bigger in one of the treatment groups
boxplot(weight ~ plot_id, data = surveys)

```