---
title: "04_manipulatingData"
author: "Jannine Chamorro"
date: "9/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lesson Plan:
-Describe the purpose of the dplyr and tidyr packages.
-Select certain columns in a data frame with the dplyr function select.
-Extract certain rows in a data frame according to logical (boolean) conditions with the dplyr function filter .
-Link the output of one dplyr function to the input of another function with the ‘pipe’ operator %>%.
-Add new columns to a data frame that are functions of existing columns with mutate.
-Use the split-apply-combine concept for data analysis.
-Use summarize, group_by, and count to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.

#1. Reading In and Inspecting Data

```{r}
## reading in our data


## inspect the data


## preview the data


```

#2. Selecting Columns and Filtering Rows

Next, we’re going to learn some of the most common dplyr functions:

select(): subset columns
filter(): subset rows on conditions
mutate(): create new columns by using information from other columns
group_by() and summarize(): create summary statistics on grouped data
count(): count discrete values

```{r}
## to select colums we use select()



## to select ALL columns EXCEPT certain ones, put a “-” in front of the variable to exclude it



## to choose rows based on a specific criterion, use filter()

```

#3. Pipes
```{r}
## selecting and filtering at the same time, using intermediate steps
## create a new dataframe (df) by filtering our orginial df



## now we can select columns from our new df




## selecting and filtering by nesting functions




## selecting and filtering using pipes
##  PIPES ( %>% ) let you take the output of one function and send it directly to the next, which is useful when you need to do many things to the same dataset.


## creating a new object/df with this smaller version of the data

```


#4. Mutate
```{r}
## create new columns based on the values in existing columns

## create and column of weight in kg



## create a new column based on the first new column 



## create a new column based on the first new column 
## just looking at the first few rows



## remove NAs using filter function

```

#5. Summarize

split-apply-combine paradigm: split the data into groups, apply some analysis to each group, and then combine the results

```{r}
## compute mean weight by sex


## group by multiple columns



## remove the missing values for weight before we attempt to calculate the summary statistics on weight
## because the missing values are removed first, we can omit na.rm = TRUE when computing the mean:


## printing a specific amount of rows



## we could add a column indicating the minimum weight for each species for each


## counting



## if we wanted to count the number of rows of data for each sex




## grouping by a variable, and summarizing it by counting the number of observations in that group



## Previous example shows the use of count() to count the number of rows/observations for one factor (i.e., sex).
## If we wanted to count combination of factors, such as sex and species, we would specify the first and the second factor as the arguments of count():

```

