---
title: "04_manipulatingData_WITHCODE"
author: "Jannine Chamorro"
date: "9/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lesson Plan:
-Describe the purpose of the dplyr and tidyr packages.
-Select certain columns in a data frame with the dplyr function select.
-Extract certain rows in a data frame according to logical (boolean) conditions with the dplyr function filter .
-Link the output of one dplyr function to the input of another function with the ‘pipe’ operator %>%.
-Add new columns to a data frame that are functions of existing columns with mutate.
-Use the split-apply-combine concept for data analysis.
-Use summarize, group_by, and count to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.

#1. Reading In and Inspecting Data

```{r}
## load packages
library(tidyverse)

## reading in our data
data = read_csv("surveys.csv") # remember to check your working directory!

## inspect the data
str(data)
dim(data)
names(data)

## preview the data
View(data)

```

#2. Selecting Columns and Filtering Rows

Next, we’re going to learn some of the most common dplyr functions:

select(): subset columns
filter(): subset rows on conditions
mutate(): create new columns by using information from other columns
group_by() and summarize(): create summary statistics on grouped data
count(): count discrete values

```{r}
## to select colums we use select()

select(data, month, day, plot_id)

## to select ALL columns EXCEPT certain ones, put a “-” in front of the variable to exclude it

select(data, -month, -day, -year)

## to choose rows based on a specific criterion, use filter()

filter(data, sex == "M")
filter(data, sex =="M" & species_id == "NL")
```

#3. Pipes
```{r}
## selecting and filtering at the same time, using intermediate steps
## create a new dataframe (df) by filtering our orginial df

df =  filter(data, species_id == "NL")

## now we can select columns from our new df

select(df, year, month, day)

## selecting and filtering by nesting functions

select(filter(data, species_id == "NL"),
       year, month, day) #this should give you the same result as the two lines above


## selecting and filtering using pipes
##  PIPES ( %>% ) let you take the output of one function and send it directly to the next, which is useful when you need to do many things to the same dataset.

df = data %>% 
  filter(species_id == "NL") %>% 
  select(year, month, day)

## creating a new object/df with this smaller version of the data

str(df)
```


#4. Mutate
```{r}
## create new columns based on the values in existing columns

## create and column of weight in kg from current weight (g)

data_2 = data %>% 
  mutate(weight_kg = weight/1000)

## create a new column based on the first new column 
#length-weight ratio

data_2 = data %>% 
  mutate(weight_kg = weight/1000,
         lwr = hindfoot_length / weight_kg)


## create a new column based on the first new column 
## just looking at the first few rows
head(data_2)


## remove NAs using filter function

data_3 = data_2 %>% 
  filter(!is.na(lwr))

#another option:
data_3 = data_2 %>% 
  na.omit() #what is the difference between these two??
```

#5. Summarize

split-apply-combine paradigm: split the data into groups, apply some analysis to each group, and then combine the results

```{r}
## compute mean weight by sex

data_2 %>% 
  group_by(sex) %>% 
  summarise(mean_weight = mean(weight, na.rm = T))

## group by multiple columns

data_2 %>% 
  group_by(sex,species_id) %>% 
  summarise(mean_weight = mean(weight, na.rm = T))

## remove the missing values for weight before we attempt to calculate the summary statistics on weight
## because the missing values are removed first, we can omit na.rm = TRUE when computing the mean:
data_2 %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight))

## printing a specific amount of rows
data_2 %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight)) %>% 
  head(6)


## we could add a column indicating the minimum weight for each species for each

data_2 %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))

## counting



## if we wanted to count the number of rows of data for each sex

data_2 %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight),
            min_weight = min(weight),
            n = n())


## grouping by a variable, and summarizing it by counting the number of observations in that group

data_2 %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  count()

## Previous example shows the use of count() to count the number of rows/observations for one factor (i.e., sex).
## If we wanted to count combination of factors, such as sex and species, we would specify the first and the second factor as the arguments of count():

```

